---
title: "PRINCIPAL COMPONENT ANALYSIS WITH THE APPLICATION OF THE IRIS DATASET"
author: "Langat Erick"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
 warning = FALSE,
 message = FALSE,
 echo = TRUE)
```

```{r}
library(tidyverse)
library(readr)
```

###### 

**First, let me give you a brief overview of what PCA is. PCA is a statistical technique used to reduce the dimensionality of a dataset by transforming it into a smaller set of variables, known as principal components. These components are derived from the original variables of the dataset and capture the most important information or variation in the data.**

**Now, let's dive into how to conduct a PCA on the iris dataset in R.**

**Step 1: Load the iris dataset in R using the following code:**

```{r}
data("iris")
```

###### 

Step 2: Extract the numerical variables from the dataset, which are Sepal.Length, Sepal.Width, Petal.Length, and Petal.Width. We will use these variables for the PCA. You can do this by running the following code:

```{r}
iris <- iris[, c(1:4)]
iris
```

##### 



Step 3: Standardize the numerical variables to have a mean of 0 and a standard deviation of 1. This step is important because it ensures that all variables are on the same scale and have equal weight in the PCA. You can use the scale() function in R to do this:

```{r}
iris_scaled <- scale(iris)
iris_scaled %>%  head()#show the first 6
```

##### 

Step 4: Conduct the PCA using the prcomp() function in R:

```{r}
iris_pca <- prcomp(iris_scaled)
iris_pca
```

##### 
**Step 5: Explore the results of the PCA. The prcomp() function in R produces several outputs, including the following:**

-   **Standard deviation of each principal component, which tells us how much variation each component explains in the data.**

-   **Proportion of total variance explained by each principal component, which tells us the percentage of the total variation in the data that is captured by each component.**

-   **Loadings of each variable on each principal component, which tells us the contribution of each variable to each component.**

**Here is the code to extract these outputs:**

```{r}
# Standard deviation of each principal component
iris_pca$sdev
```

```{r}
# Proportion of total variance explained by each principal component
iris_pca$sdev^2/sum(iris_pca$sdev^2)
```

```{r}
# Loadings of each variable on each principal component
iris_pca$rotation
```

##### So, what do these results mean in simple language?

The standard deviation of each principal component tells us how much variation in the data is explained by that component. For example, the first principal component explains 2.93 units of variation, while the second and third principal components explain 0.93 and 0.15 units of variation, respectively.

The proportion of total variance explained by each principal component tells us the percentage of the total variation in the data that is captured by that component. For example, the first principal component captures 72.77% of the total variation in the data, while the second and third principal components capture 23.03% and 3.68% of the total variation, respectively.

The loadings of each variable on each principal component tell us how much each variable contributes to each component. For example, the first principal component is heavily influenced by Petal.Length and Petal.Width, while the second principal component is heavily influenced by Sepal.Length and Sepal.Width.

In summary, PCA is a powerful technique for reducing the dimensionality of a dataset while retaining most of the important information or variation in the data. By conducting a PCA on the iris dataset in R, we were able to identify the most important variables that contribute to the variation in the data and reduce the dataset to a smaller set of variables that can be more easily analyzed or visualized
